---
title: "Convergence in lme4 and lme4.0"
author: "Phillip Alday"
date: `r as.character(Sys.time())`
output: html_document
---

```{r pkgs,message=FALSE}
library(knitcitations)
library(RCurl)
library(lme4)
library(lme4.0)
lme4ver <- as.character(packageVersion("lme4"))
lme4.0ver <- as.character(packageVersion("lme4.0"))
```

```{r opts,echo=FALSE}
opts_chunk$set(tidy=FALSE)
```

The differences in convergence between "old" and "new" `lme4` is well known and often leads to substantially poorer fits with new lme4 for psycholinguistic datasets `r citep("https://hlplab.wordpress.com/2014/03/17/old-and-new-lme4/")`. For the data set here, the difference in fits leads to similar estimates but very different standard errors and thus $t$-values. 

The following was done using lme4 `r lme4ver` and lme4.0 `r lme4.0ver`.

```{r setup}
c. <- function(x) scale(x, center=TRUE, scale=FALSE)
cs. <- function(x) scale(x, center=TRUE, scale=TRUE)
reml <- FALSE
modeldata <- read.table("data.tab",header=TRUE,
                        colClasses=c(mean="numeric",
                        subj="factor",item="factor",
                        roi="factor",win="factor",
                        sdiff="numeric", dist="numeric",
                        signdist="numeric",ambiguity="factor"))

modeldata <- subset(modeldata, roi == 'Left-Posterior')
modeldata.n400 <- subset(modeldata,win=="N400")
form <- mean ~ ambiguity * c.(sdiff) + (1+c.(sdiff)|item) +
    (1+c.(sdiff)|subj)
form.s <- mean ~ ambiguity * cs.(sdiff) + (1+cs.(sdiff)|item) +
    (1+cs.(sdiff)|subj)
```

## Plots

(Haven't found a great way to do this, just trying to look
to see whether there is something odd about the data ...)

```{r pix}
library(ggplot2); theme_set(theme_bw())
mm <- transform(modeldata.n400,sdiff=cs.(sdiff),
                subj=reorder(subj,mean),
                item=reorder(item,mean))
mm0 <- subset(mm,select=c(mean,sdiff,ambiguity,subj,item))
## save("mm0",file="mm0.RData")
## L <- load("mm0.RData")
ggplot(mm0,aes(mean,subj,colour=sdiff))+geom_point(alpha=0.5)+
    facet_grid(ambiguity~.)
```

# lme4

```{r lme4fitnew,cache=TRUE}
t.new <- system.time(sdiff.new <- lme4::lmer(form,
                                             data=modeldata.n400, REML=reml))
t.new
print(summary(sdiff.new),correlation=FALSE)
```
Note we still get convergence warnings, even with lme4 `r lme4ver`; this means they are probably **not** just false positives (so in this case they are a good thing!)  Should have gotten scaling warnings too ???

```{r lme4fitnew.S,cache=TRUE}
sdiff.new.scaled <- update(sdiff.new,formula=form.s)
```

# lme4.0
```{r lme4fitold,cache=TRUE}
t.old <- system.time(sdiff.old <- lme4.0::lmer(form,
                                               data=modeldata.n400, REML=reml))
t.old
summary(sdiff.old)
```

As stated, the important difference in the fixed effects is the
size of the standard error in `c.(sdiff)` ...

```{r coefplot,echo=FALSE}
tf <- function(x,model="") {
    x2 <- as.data.frame(coef(summary(x)))
    x2$var <- rownames(x2)
    rownames(x2) <- NULL
    x2 <- x2[-1,c(4,1,2)]
    names(x2) <- c("var","est","stder")
    data.frame(model,x2)
}
cc <- rbind(tf(sdiff.new,"new"),tf(sdiff.old,"old"))
ggplot(cc,aes(x=model,y=est,ymin=est-2*stder,ymax=est+2*stder,
        colour=model))+geom_pointrange()+
    facet_wrap(~var,scale="free")
```

Compare theta values? (We can't get confidence intervals without
profiling ...)


# lme4 starting from lme4.0 results
```{r lme4fitstart,cache=TRUE}
oldtheta <- getME(sdiff.old,"theta")
newtheta <- lme4::getME(sdiff.new,"theta")
sdiff.oldstart <- lme4::lmer(form,
                             data=modeldata.n400, REML=reml,
                             start=list(theta=getME(sdiff.old,"theta")))
sdiff.oldstart.nlminb <- lme4::lmer(form,
                             data=modeldata.n400, REML=reml,
                             start=list(theta=getME(sdiff.old,"theta")),
                             control=lmerControl(optimizer="optimx",
                             optCtrl=list(method="nlminb")))
```

We have an interesting problem here.  New `lme4` calculates the
deviance slightly differently; at $\hat \theta_{\textrm{old}}$
it computes a deviance value about 29 units higher than `lme4.0`.
If we start `lme4` from that point, we actually get a **worse**
result (the deviance goes up by about 480 units!) --- presumably
we get thrown off by the initial displacement of the points from
the starting value ...

```{r cmpdevs}
sdiff.newdev <- update(sdiff.new,devFunOnly=TRUE)
devvec <- c(old=deviance(sdiff.old),new=deviance(sdiff.new),
            oldtheta.newdev=sdiff.newdev(oldtheta),
            oldstart=deviance(sdiff.oldstart),
            oldstart.nlminb=deviance(sdiff.oldstart.nlminb))
print(sort(devvec-min(devvec)),digits=3)
```


```{r get_allfit}
afurl <- paste0("https://raw.githubusercontent.com/lme4/",
                 "lme4/master/misc/issues/allFit.R")
eval(parse(text=getURL(afurl)))
```

I should suppress the warnings here (they'll be preserved in the `@optinfo` slot anyway), but that will require refreshing the cache, so I'm not going to do it right now ...)  The second `allFit` call (starting from the `lme4.0` fitted values) doesn't work at all, for some structural reason I don't understand (having to do with package loading/unloading I think).  All of this is sufficiently lengthy that I should consider putting it in a batch file rather than relying on knitr caching ...

```{r comp_allfit,warning=FALSE,message=FALSE,cache=TRUE}
detach("package:lme4.0")
t.all <- system.time(sdiff.new.all <- allFit(sdiff.new))
sdiff.oldstart.all <- allFit(sdiff.oldstart)
library(lme4.0)
```

The bottom line is that trying lots of different optimizers doesn't do us any good in this case. Oddly enough, our Nelder-Mead implementation (which we have found to be generally less reliable than BOBYQA for big problems) turns out to do *best* of all the options (other than `lme4.0`) in this case ...

```{r cmplik}
is.OK <- !sapply(sdiff.new.all,inherits,"error")
all.dev <- c(sapply(sdiff.new.all[is.OK],deviance),lme4.0=deviance(sdiff.old))
print(sort(all.dev-min(all.dev)),digits=4)
```

Now try with scaled data:
```{r comp_allfit_scaled,warning=FALSE,message=FALSE,cache=TRUE}
detach("package:lme4.0")
t.all <- system.time(sdiff.new.scaled.all <- allFit(sdiff.new.scaled))
library(lme4.0)
```

The code and data for this example can be found on [GitHub](https://github.com/palday/lme4-convergence)

